\textbf{Hardware accelerator coupling in heterogeneous systems: an evaluation in the security domain}

### 2019-12-09 reports/paper.tex
write more, almost done

\title{\textbf{Hardware accelerator topologies in heterogeneous systems: an evaluation in the
    security domain}}
\author{Bruno Morais}
\date{}

@@

\begin{abstract}
  Hardware accelerators (HWACCs) have driven performance gains for
  domain specific applications which leverage highly optimized computation
  cores. However, the traditional accelerator access topology as an add-on to
  general purpose processors through shared memory space/bus, reveals performance
  issues as the number of HWACCs grow and systems become
  communication-bound.
  On the other hand, efficient development and research of HWACCs
  that are closer to a general purpose processor's datapath is seen as a
  prohibitive task due to the required microarchitecture knowledge and lack of
  tools to explore the design space. This has in turn hindered the exploration of
  such architectures for gains in either performance or security spectrums.
  RISC-V projects made available in the last few years provide a framework that
  greatly eases microarchitecture research, enabling further experimentation
  into security and performance aspects for heterogeneous platforms.
  This work uses a customized RISC-V SoC to explore performance trade-offs
  between different HWACC topologies (in-datapath, memory-mapped) for embedded
  systems in the security domain by implementing an IP core for cryptographic
  primitive acceleration. We also show a hybrid accelerator topology for
  acceleration and/or secure memory storage which offers a speedup of 6x compared to a
  pure software implementation.
  % results also go in here

@@

\begin{figure}
  \centering
  \include{media/loosely_acc}
  \caption{Loosely-coupled HWACCs: typical memory mapped HWACC usage in a heterogenous architecture.}
\end{figure}

\begin{figure}
  \centering
  \resizebox{0.3\textwidth}{!}{
    \begin{minipage}{0.5\textwidth}
      \centering
      \include{media/tightly_acc}
    \end{minipage}
  }
  \caption{Tightly-coupled HWACCs: accelerator is moved into CPU datapath,
    operating similarly to microarchitectural Functional Units (FUs).}
\end{figure}

@@

\subsection{Hardware accelerators?}
% Tightly vs loosely coupled hwaccs

\subsection{RISC-V instruction set extensions}
% Xcrypto

@@

Synthesis and implementation are performed using Vivado. This is only done
once, for the MMIO accelerator, as the core implementation of Simon is shared
across all accelerators. Usage as reported by vivado is at 3119 LUTs and 2004
flip flops, which account for approximately 17\% and 19.5\% of the total design
(SoC) size.

@@

Estimated power consumption is also reported after synthesis and implementation
using Vivado. It is reported as 8mW for the simon accelerator alone --- the
entire SoC power consumption is estimated at 252 mW (HWACC accounts for \approx 3\%).

>>>>

\title{\textbf{Hardware accelerator coupling in heterogeneous systems: an evaluation in the
    security domain}}
\author{Bruno Morais \texttt{<soutomaiormunizmo.b@husky.neu.edu>}}
\date{}

@@

\begin{abstract}
  Hardware accelerators (HWACCs) have driven performance gains for
  domain specific applications which leverage highly optimized computation
  cores. However, the traditional accelerator access topology as an add-on to
  general purpose processors through shared memory space/bus, reveals performance
  issues as the number of HWACCs grow and systems become
  communication-bound.
  On the other hand, efficient development and research of HWACCs
  that are closer to a general purpose processor's datapath is seen as a
  prohibitive task due to the required microarchitecture knowledge and lack of
  tools to explore the design space. This has in turn hindered the exploration of
  such architectures for gains in either performance or security spectrums.
  RISC-V projects made available in the last few years provide a framework that
  greatly eases microarchitecture research, enabling further experimentation
  into security and performance aspects for heterogeneous platforms.
  This work uses a customized RISC-V SoC to explore performance trade-offs
  between different HWACC topologies (in-datapath, memory-mapped) for embedded
  systems in the security domain by implementing an IP core for cryptographic
  primitive acceleration. We also show a hybrid accelerator topology for
  acceleration and/or secure memory storage which offers a speedup of 6x compared to a
  pure software implementation.

@@

The demand for power-efficient computing, specially on edge and/or embedded
devices has continuously driven the development of hardware accelerators
(HWACCs) to meet these power requirements. General-purpose computing units alone
are too flexible and cannot achieve the same parallelism as a specialized
accelerator. The growing complexity of algorithms, specially with the advent of
widespread machine learning and security needs such as authentication, integrity
verification and privacy, has pushed the boundaries of embedded systems,
dramatically increasing the number of accelerators present in sophisticated
embedded SoCs.

With the number of HWACCs rapidly increasing, the use of typical memory mapped HWACC,
which shares the memory bus of a general purpose CPU creates bottlenecks~\cite{TSS},
specially for certain classes of applications, for instance streaming-based
applications, where communication between HWACCs (function compositions) is a requirement.
Figure~\ref{fig:loosely} shows the typical memory-mapped HWACC, also called a
\textit{loosely-coupled} HWACC.

\begin{figure}
  \centering
  \include{media/loosely_acc}
  \caption{Loosely-coupled HWACCs: typical memory mapped HWACC usage in a heterogenous architecture.}
  \label{fig:loosely}
\end{figure}

In constrast with the \textit{loosely-coupled} accelerator which is external to
the CPU datapath, an alternative topology is to bring the HWACC much closer, and
effectively into the CPU datapath itself. Those are called
\textit{tightly-coupled} HWACCs. Essentially, in this configuration, the HWACC
can be seen as one of the CPU's functional units, dedicated to a certain
specific task. Figure~\ref{fig:tightly} shows a simplified CPU datapath with
tightly-coupled accelerators.

While it is relatively straightforward to design accelerators and perform research using
loosely-coupled accelerators by deploying them in FPGAs, for instance, research
into tightly-coupled HWACCs has been historically prohibitive due to the need to
have an easily modifiable processor implementation and all the necessary
supporting tools such as compiler toolchain, simulators and a synthesis enabled
design. The RISC-V project and its community have fostered the development of
such a system, which is open source and user friendly, enabling much easier
research of such a topology.

Tightly-coupled accelerators are invaluable from a security perspective, not
only to accelerate cryptographic functions but also to enable secure data
storage. The main contribution of this work is the implementation of a HWACC
that accelerates the Simon cipher and an analysis of the speedup provided by
different couplings. We also show a hybrid tightly-coupled accelerator which has
access to the on-chip coherent memory hierarchy and enables secure data storage
and retrieval to/from main memory.

\begin{figure}
  \centering
  \resizebox{0.3\textwidth}{!}{
    \begin{minipage}{0.5\textwidth}
      \centering
      \include{media/tightly_acc}
    \end{minipage}
  }
  \caption{Tightly-coupled HWACCs: accelerator is moved into CPU datapath,
    operating similarly to microarchitectural Functional Units (FUs).}
  \label{fig:tightly}
\end{figure}

@@

\subsection{Hardware accelerator coupling in heterogeneous systems}
The usage of different couplings (tightly/loosely coupled) leads to mixed
results depending on the domain of the application. Typical usage of HWACCs fall
within the loosely-coupled category, which pays little attention to interaction
with on-chip components such as caches~\cite{tightLoose}.

There is potential gain by moving accelerators into the CPU datapath and
enabling leveraging the memory hierarchy to increase speedup. This incurs in
more development costs, as discussed before when compared to loosely-coupled
accelerators, which are more easily abstracted and do not need extensive
modifications to the toolchain.

\subsection{RISC-V cryptographic extensions}
The RISC-V community has formed a group to specify ISA extensions for
cryptographic primitives~\cite{RiscVCrypto}. The approach is similar to Intel's
AES-NI~\cite{AESNI}, and is expected to support at least AES and SHA-2. These
type of accelerators are tightly-coupled accelerators directly controlled by
software instructions.

There is also some independent work showcasing the implementation of accelerated
cryptographic functions by means of instruction set extensions by the University
of Bristol~\cite{xcrypto}. The XCrypto is part of a larger project and
implements AES, SHA-2 and SHA-3.

@@

Synthesis and implementation are performed using Vivado. This is only done
once, for the MMIO accelerator, as the core implementation of Simon is shared
across all accelerators. Usage as reported by vivado is at 3119 LUTs and 2004
flip flops, which account for approximately 17\% and 19.5\% of the total design
(SoC) size. The target FPGA has a total of 20800 and 41600 LUTs and flip flops
available, respectively.

@@

Estimated power consumption is also reported after synthesis and implementation
using Vivado. It is reported as 8mW for the simon accelerator alone --- the
entire SoC power consumption is estimated at 252 mW (HWACC accounts for $\approx$ 3\%).
